Introducing an Ecological Trait-data Standard

### 2017-11-08 manuscript/manuscript.Rmd
updates on manuscript

---
title: "Introducing an Ecological Trait-data Standard"
author: Florian D. Schneider, Malte Jochum, Gaetane LeProvost, Andreas Ostrowski,
  Caterina Penone, Nadja K. Simons
csl: amnat.csl
output:
  word_document:
    keep_md: yes
  pdf_document: default
geometry: margin=1.5in
header-includes: \usepackage{lineno}
fontsize: 12pt
bibliography: div05.bib
---

@@

Given the expected future challenges in dealing with climate change and managing land use, such global trait synthesis is an interdisciplinary/cross-topic field with great potential. 
<!-- Prospect: what will trait analysis acheive, if global data across taxa are available, -->
<!-- Nadja: Some thoughts/snippets based on other publications: " a relevant amount of data is available to quantify and understand trait variation beyond aggregation on species level. The analysis presented here is only a first step to disentangle within- and between-species variability. [...] A more empirically based representation of functional diversity is expected to contribute to an improved prediction of biome boundary shifts in a changing environment. [...] Empirical inference of this kind is expected to stimulate and strengthen hypothesis-driven research (Golub, 2010; Weinberg, 2010) towards a unified ecological assessment of plant traits and their role for the functioning of the terrestrial biosphere." [@kattge11] / "Our traitbased approach to modelling competition makes generalization possible across the forest ecosystems of the world and their highly diverse species composition. [...] A challenge for the future is to move beyond growth to analyse all key demographic rates and life-history stages, and analyse how traits influence competitive outcomes and stable coexistence at the population level."[@Kunstler2016] / "Traits include those that in combination allow a first-order assessment of species' similarity in "ecological niches" or the overlap in their functional role. The standardized interpretation and adherence to source information and clear documentation aims to provide a replicable and transparent basis for research. [...] The trait variables in the dataset should support hypothesis testing or applied inference in ecology, conservation and evolution." [@Wilman2014] -->

@@

Open-access publications in low-threshold data repositories such as Figshare, dataDryad, Researchgate or Zenodo.org are gaining importance and foster a decentralised data hosting with low expectations regarding data standardization and documentation. Although open for manual download, data on those platforms might be stored in proprietary (e.g. .xlsx) or binary (e.g. .pdf) data formats which make a programmatical extraction tedious or impossible. Since the platforms are open for any type of scientific data, no data standards are enforced upon the uploaded material. Given this prospect, there will be no lack of data, but a lack of data standardization hampering efficient future use of published datasets.

@@

Ontologies have been developed for plant traits, for instance in the TOP Thesaurus of plant traits, which is employed in the TRY database [@garnier17], the Planteome ontology for crop plants [@walls12; @jaiswal05; www.plantontology.org], or the Flora Phenotype Ontology [http://obofoundry.org/ontology/flopo.html @hoehndorf16]. On the side of animal traits, few initiatives published comprehensive trait standards and method catalogues, like for instance a Thesaurus for Soil Invertebrate Trait-based Approaches [T-SITA, http://t-sita.cesab.org/, @pey14] or the vertebrate trait ontology [http://obofoundry.org/ontology/vt.html; @park13]. A resource for biological ontologies and terminologies is the OBO foundry (http://obofoundry.org/).

While these ontologies define traits for focal groups of organisms, they do not specify in which formats or structures trait data should be stored. 
The Encyclopedia of Life (EOL) project has advanced on the topic by developing the database framework TraitBank [@parr15], which is to date the most general approach to providing a semantic structure for trait data tables. The framework employs established terms provided by the Darwin Core Standard (DWC, @wieczoreck12) and relates trait definitions to specific trait ontologies and phenotypic and anatomical ontologies (<!-- define --> PATO, UBERON), and maps taxa to global identifiers in taxonomic hierarchies of multiple name service providers (e.g. GBIF Taxonomy) to capture synonyms, misspellings and controversies  [@parr15]. Additional layers of information capture bibliographic reference, multimedia archives and ecological interactions. TraitBank invites data submissions to the EOL database in a structured Darwin Core Archive. 
Another initiative aimed for a mobilization of measurements from museal collection data collated in the VertNet database into harmonized traitdata [@guralnick16].

However, the Terms of the Darwin Core Standard and available Extensions, TraitBank and VertNet, are responses to a demand in Biodiversity Data Management for structured input from the research communities. The practical importance of these frameworks for data providers and data users is less obvious and often not appropriately considered. To date, the existing trait data standards do not cover all of the aspects required in the various use-cases of trait-based research. <!-- examples for blind spots -->

@@

With this paper, we propose a roadmap towards a comprehensive Ecological Traitdata Standard (EST) for general application in trait-based research, particularly building upon the work of @parr16, @kattge11, and @garnier17, and drawing on the combined expertise of empirical biodiversity researchers (data providers), biodiversity synthesis researchers (data users), and biodiversity informatics researchers (data managers).

We define a versatile, minimal data structure, as well as a scheme of defined column names that captures the different degrees of resolution and measurement detail for multiple use cases of trait data. We also provide computational tools in an R package to help users transfer their original data into the proposed scheme.

With this framework, we build upon the structure proposed for EOL TraitBank [@parr16], which applies and expands terms of the Darwin Core standard [DWC; @wieczorek12]. We further expand the definitions of these terms for the use case of trait data and add terms which cover the special demands of trait-based research, while maintaining semantic compatibility with existing initatives.

@@

As a minimum consensus, trait datasets may be defined as follows: 
A trait-dataset contains quantitative measurements or qualitative facts (i.e. trait values) about physical phenotypic characteristics of fitness, behaviour or performance (i.e. traits) observed on an individual or parts of an individual (i.e. an entity) that can be assigned to a biological taxon (i.e. a species). The entity or observation (i.e. the occurrence) to which the reported measurement or fact applies may differ in resolution -- depending on the scientific question -- and could be a subsample or bodypart, an individual specimen, an entire species or higher-level taxon.  
This definition refines the entity-quality model (EQ) as described by @garnier17, where a trait observation is phrased as 'an entity having a quality'.

<!-- Cat: the difference [between quantitative measurements or qualitative facts] is not totally clear to me - is it the difference between field vs literature gathered data?  Flo: I see measurements as quantitative and facts as qualitative traits, which mostly aligns with field vs. literature, but not entirely. A differentiation that has rarely been specified in the trait-litereature it seems. Garnier et al 2017 just speak of "quality", which is assigned to an "entity", without specifiying it further. Nadja: I like this differentiation, to me this nicely reflects morphometric and literature traits (although its of course not exclusive). Maybe it helps if we add an example after the definition?-->

Trait-datasets usually take different formats. For instance, if trait data have been collated at the species level from different literature sources or from expert knowledge, or as aggregated measurements collated from raw data, they usually are reported in a species $\times$ trait matrix (a.k.a. wide-table format), with a column of trait values for each trait recorded and a row for each species (or taxon) for which data were available. This format is usually reporting missing data as NA. It may store additional information (e.g. on variation of means or literature source) in secondary colums or accompanying datasheets. 
The matrix format is widely used for the production of lookup-tables at the species level, which may either report qualitative facts (from literature or  expert knowledge) or aggregated quantitative traits (averages or ranges for this taxon). Quantitative traits at this resolution for instance may be used for the calculation of community weighted means or functional diversity metrics at the community level [@laliberte10a]. Also at this level of information, missing entries, e.g. about the trait values or behaviour and functional role of a rarely described species, may be inferred from these tables combined with phylogenies ['gap filling', @swenson14] and using bayesian approaches [@schrodt15]. However, the tables do not allow to estimate within-species variation or assess the quality of the reported measurement or fact [@guralnik15].

Phenotypic variation of traits within species would usually be recorded per observation, i.e. each individual specimen would be recorded in a row, with the respective trait values stored in columns. In these occurrence data tables, each row is centered around a single physical occurrence of a species. Those data are common in investigations of evolutionary trade-offs and trait correlation, or of intra-specific variation along environmental gradients. Occurrence data would also capture phenotypic variation arising from morphotypes and sexual dimorphism. This format is the most intuitive for recording own empirical measurements and therefore is common for measured quantitative data and rarely found for reported qualitative facts obtained from the literature.

Computationally more effective and allowing for highest flexibility is the storage of traitdata in 'tidied' long-table formats [@wickham14a], where each row is reserved for a single measurement or fact of a specific trait, referenced to a single occurrence (i.e. a specimen) which is assigned to a specific taxon. This allows for storing repeated measurements on a single individual by linking the data of separate rows via a unique specimen ID. Also, multivariate trait measurements can be recorded in this format by linking multiple rows via a unique measurement ID.  Longtable dataset purport multiple advantages for the data manipulation (e.g. filtering, subsetting and aggregating data), visualization (e.g. plot measured values by factor variable or taxon) and statistical modeling (e.g. ANOVA for testing difference of trait value by sex) [@wickham14a]. They further draw from a defined set of columns, which makes it easier to merge datasets.

The latter format thus provides the highest resolution and is ideal for storing raw data of trait measurements in huge databases and for linking them via the respective unique identifiers to additional information, such as occurrenceID or locationID. Each row of the dataset can be interpreted as a statement of an 'entity x having a qualitative/quantitative feature y'. 
To comply with the EST, datasets must be reformatted to the longtable format.

@@

@@

In the subsequent sections we propose a set of terms that describe the columns composing a trait dataset, starting with the ones required for our minimal definition of 'core' trait-datasets proposed above.
We furthermore provide extensions to add information on the level of the occurrence, the measurement or fact, and the entire dataset (i.e. metadata). 
A second set of terms is intended to facilitate the definition of traits in ontologies or a thesaurus of traits, consistent with semantic web standards (@ref, see below for further detail). 
The full glossary of terms can be accessed at https://ecologicaltraitdata.github.io/TraitDataStandard/ (will be moved to GFBio Terminology Server before publication). Were possible, we relate these terms to existing terms in ontologies for biological data, like the Glossary of EOL (http://eol.org/info/516) and Darwin Core ().

A dataset that applies these terms should refer to the data standard in its metadata. Since the traitdata standard is a resource under open community development, this reference also should include the version number. This is best warranted by providing the full URI of the applied  traitdata standard in the metadata (e.g. http://terminologies.gfbio.de/traitdatastandard/v1.0 !invalid URL!).

@@

@@

Taxonomic standardisation, with respect to ongoing renamings and refinemenets, is a challenging issue in its own right. To enable compatibility across datasets, conventions on trait definitions and taxon names are inevitable. We consider the fact that most authors have their own schemes for standardisation and may refer to different scientific community standards by allowing for a double record of both user-specific and standardised entries [as practiced in TRY; @kattge11]. This dupliation of data enables continuity on the author and data owner side and quality checking and comparability on the data user side. We therefore propose to add a second set of columns that contain standardised entries of the taxon name (`scientificNameStd`) which map synonyms to an accepted species name according to a published taxonomic ontology (e.g. GBIF Backbone Terminology) and links to it using a globally unique identifier (`taxonID`). The field `taxonID` therefore should contain an public,  globally unique Uniform Resource Identifier (URI) that resolves to the precise taxon definition at the time of the description or measurement (e.g. for *Bellis perennis*, the taxonID would be  'https://www.gbif.org/species/3117424'). 
Keeping a reference to a kingdom and class (using field `kingdom`, `class` as described by Darwin Core Standard and suggested for EOL Traitbank) will avoid ambiguities with taxonomic homonyms. 
Additionally, it is best practice to keep a record of the reported taxon rank (in column `taxonRank`) for filtering purposes for traits recorded at the family or genus level. This information can be used for trait inference of higher taxonomic resolution using hierarchical probabilistic matrix factorisation [@schrodt15]. 
Alternatively to expanding the number of columns, these data could be contained in a separate data file which relates to the core dataset via the field `ScientificName`.

@@

A similar approach to standardised trait names and harmonised reported values in compliance with published ontologies is advised. Here, column `traitNameStd` should report the accepted trait name in an online ontology, while the value and unit are reported in columns `traitValueStd` and `traitUnitStd`.
Similar as for `taxonID`, the field `traitID` should refer to globally unique identifiers (URI) of trait definitions or a measurement methodology.

<!-- The initiatives to standardise traits into consensus terminology are usually formed around a specific research question, methodology, or organism group and therefore vary in the traits that they consider important.  -->
To date, semantic ontologies exist for plant traits [e.g. TOP Plant Trait Ontology , @garnier17], soil invertebrates [T-SITA, @pey14], or biomedical traits of vertebrates [Vertebrate Trait Ontology, @park13] and more use cases (see OBO foundry for further examples). However, for most organism groups and research contexts we attest a deficit of public ontologies with publicly accessible URIs.

@@

Relating the trait definition to methodological definitions on trait measurements help standardize the physical process of measurements. The Handbook for Measurement of traits [@moretti17a] is a model example of such a methodological standardization approach.

Each trait definition may link to a broader or narrower term (`broaderTerm`, `narrowerTerm`), related Terms (`relatedTerm`) or synonyms (`synonym`). For instance, the definition of 'femur length of first leg, left side' is narrower than 'femur length' which is narrower than 'leg trait' which is narrower than 'locomotion trait' (@ref semantic database methods). This extends the trait list into a semantic web resource, facilitates the classification of traits, and allows for cross-taxon comparative studies at the level of broader terms. (more detail on semantic web ontologies can be found @refs...).

The existing thesauri and ontologies differ widely in terms of hierarchical depths and detail, as well as curation efforts and measures for peer-reviewed quality control. More compatibility of trait thesauri could greatly improve the possibilities for a comparison of trait data from different sources. 
We encourage hosting such semantic web resources developed for own projects on public repositories like OBO Foundry, Github or the GFBio Terminology service. We provide an example thesaurus, that has been developed to capture traits assessed within the German Biodiversity Exploratories project on Github ().

@@

<!-- 
(We conceptualize trait-datasets to  not include information about a species' interaction partners/associations of an organism, since they are -- to our understanding -- not within the scope of functional traits. However, such network data may be of interest to be combined with trait data. ) <!-- Malte: which can only be achieved if both types of datasets have been related to an ontology of species names. --> <!--Nadja: I don't think that we need to mention this. If you think that this is important, it should be moved more to the beginning, e.g. within the section "Common terms for trait data"-->

@@

Beyond measurement units or taxon rank resolution, further information might be available that are not recorded in the raw data table, but are related to the trait type, the taxon, the individual or specimen, or to the reported fact, measurement or sampling event.

In most cases those information would be kept in seperate data sheets of a spreadsheet document, like for instance the place were a specimen has been sampled, the measurement method or statistical precision of the reported value, or the literature source from where a species value has been cited. In this case, a unique identifier might link to this other datasheet, encoding each individual occurrence of a species (`occurrenceID`), single measurements or reported facts (`measurementID`), locations of sampling (`locationID`) and sampling campaigns (`eventID`).

The trait data standard  provides two extensions of the namespace that should be used to describe these data:

@@

Many terms of this section refine terms of the Darwin Core Standard and their own [MeasurementOrFact](http://rs.tdwg.org/dwc/terms/index.htm#measureindex) and [Occurrence](http://rs.tdwg.org/dwc/terms/index.htm#Occurrence) extensions  and EOL Trait Bank use of those terms.

@@

The terms may be applied as columns in a supplementary data sheet or along with the core data described above. Having separate data sheets linked by identifiers has the advantage of tidy datasets, avoiding duplication of information (@wickham14a). They furthermore can easily be packaged into Darwin Core Archives which are ideal for structured data (@parr16, Darwin core ...).  In that case, the traitdata table should at least contain a column with the respective identifier that directs to the covariate datasheet.

However, it may be safer to add these information directly as own columns of the data table to enable an analysis of cofactors and correlations further down the road. This way, if datasets of different source are merged, the information is readily available without the risk of breaking the reference to an external datasheet.

<!-- Nadja: can those last two paragraphs be merged with the second paragraph? I feel like we are repeating ourselves here. -->

@@

Scientific taxon names are matched automatically to the GBIF Backbone Taxonomy (taxonomic ontology server) by calling the function `standardize.taxonomy()`

The `standardize.traits()` function matches user-provided trait names onto public trait ontologies (for now using GBIF Backbone Taxonomy) and links it to a trait ontology or a user-provided table of trait definitions, i.e. a thesaurus of traits. The same function harmonises trait values into target units and legit factor levels.

@@

Since the intention of the package is to facilitate upload of own data to hosting services but also simplify the access to published trait data, the package includes data objects that are compiled from datasets listed in our online spreadsheet that have been published in the public domain [@gossner15; @gossner15c; @ricklefs17; @vanderplas17]. 
To access data from public data-bases, the R-package 'traits' [@chamberlain17] contains functions to extract trait data via several open API interfaces including Birdlife, EOL TraitBank or BetyDB. The package  'TR8' provides similar access to plant traits from a list of databases (including LEDA, BiolFlor and Ellenberg Values)[@bocci15] and aggregates them into a species-trait matrix. However, none of these packages provides harmonization of traits into a unified scheme to facilitate further analysis.  
We provide a script (see online appendix and R-package vignette) to demonstrate the harmonization of these different data sources into our data standard.

>>>>

---
title: "Introducing an Ecological Trait-data Standard"
author: Florian D. Schneider, Malte Jochum, Gaetane LeProvost, Andreas Ostrowski,
  Caterina Penone, Nadja K. Simons
csl: amnat.csl
output:
  pdf_document: default
  word_document:
    keep_md: yes
geometry: margin=1.5in
header-includes: \usepackage{lineno}
fontsize: 12pt
bibliography: div05.bib
---

@@

Given the expected future challenges in dealing with climate change and managing land use, such global trait synthesis is an interdisciplinary/cross-topic field with great potential. 
<!-- Nadja: Some thoughts/snippets based on other publications: " a relevant amount of data is available to quantify and understand trait variation beyond aggregation on species level. The analysis presented here is only a first step to disentangle within- and between-species variability. [...] A more empirically based representation of functional diversity is expected to contribute to an improved prediction of biome boundary shifts in a changing environment. [...] Empirical inference of this kind is expected to stimulate and strengthen hypothesis-driven research (Golub, 2010; Weinberg, 2010) towards a unified ecological assessment of plant traits and their role for the functioning of the terrestrial biosphere." [@kattge11] / "Our traitbased approach to modelling competition makes generalization possible across the forest ecosystems of the world and their highly diverse species composition. [...] A challenge for the future is to move beyond growth to analyse all key demographic rates and life-history stages, and analyse how traits influence competitive outcomes and stable coexistence at the population level."[@Kunstler2016] / "Traits include those that in combination allow a first-order assessment of species' similarity in "ecological niches" or the overlap in their functional role. The standardized interpretation and adherence to source information and clear documentation aims to provide a replicable and transparent basis for research. [...] The trait variables in the dataset should support hypothesis testing or applied inference in ecology, conservation and evolution." [@Wilman2014] -->
<!-- Prospect: what will trait analysis acheive, if global data across taxa are available,

- intra-specific variation and evolutionary trade-offs
- global variation
- community functional pattern, comparative study across continents
- big data challenges
-->

@@

Open-access publications in low-threshold data repositories such as Figshare, dataDryad, Researchgate or Zenodo.org are gaining importance and foster a decentralised data hosting with low expectations regarding data standardisation and documentation. Although open for manual download, data on those platforms might be stored in proprietary (e.g. .xlsx) or binary (e.g. .pdf) data formats which make a programmatical extraction tedious or impossible. Since the platforms are open for any type of scientific data, no data standards are forced upon the uploaded material. 
Wheter open or not, the publication of original data is increasingly required by journals to warrant reproducibility of science. Given this prospect, there will be no lack of data, but a lack of data standardization hampering efficient future use of published datasets.

@@

Ontologies have been developed for plant traits, for instance in the TOP Thesaurus of plant traits, which is employed in the TRY database [@garnier17], the Planteome ontology for crop plants [@walls12; @jaiswal05; www.plantontology.org], or the Flora Phenotype Ontology [http://obofoundry.org/ontology/flopo.html @hoehndorf16]. On the side of animal traits, a couple of initiatives published comprehensive trait standards and method catalogues, like for instance a Thesaurus for Soil Invertebrate Trait-based Approaches [T-SITA, http://t-sita.cesab.org/, @pey14] or the vertebrate trait ontology [http://obofoundry.org/ontology/vt.html; @park13]. A resource for biological ontologies and terminologies is the OBO foundry (http://obofoundry.org/).

While these ontologies define traits for focal groups of organisms, they do not specify in which formats or structures trait data should be stored. 
The Encyclopedia of Life (EOL) project has advanced on the topic by developing the database framework TraitBank [@parr15], which is to date the most general approach to providing a semantic structure for trait data tables. The framework employs established terms provided by the Darwin Core Standard (DWC, @wieczoreck12), relates trait definitions to specific trait ontologies and phenotypic and anatomical ontologies (<!-- define --> PATO, UBERON), and maps taxa to global identifiers in taxonomic hierarchies of multiple name service providers (e.g. GBIF Taxonomy) to capture synonyms, misspellings and controversies [@parr15]. Additional layers of information capture bibliographic reference, multimedia archives and ecological interactions. TraitBank invites data submissions to the EOL database in a structured Darwin Core Archive. 
Another initiative aimed for a mobilization of measurements from museal collection data collated in the VertNet database into harmonized traitdata [@guralnick16].

However, the Terms of the Darwin Core Standard and available Extensions, TraitBank and VertNet, are responses to a demand in Biodiversity Data Management for structured input from the research communities. The practical importance of these frameworks for data providers and data users is less obvious and often not appropriately considered when storing or uploading data at file-publishing services. None of the aggregated and harmonized databases for birds, amphibians, or mammals employs such ontologies or semantic web standads. One reason certainly is compatibility: to date, the existing trait data standards do not cover all of the aspects required in the various use-cases of trait-based research. <!-- examples for blind spots --> 
The other reason is competence: trait-data providers are not trained in the demands of biodiversity data-management and complying with the apparently non-intuitive data structures is an investment without immediate pay-off.

@@

With this paper, we propose a versatile, minimal data structure, as well as a scheme of a defined vocabulary that captures the different degrees of resolution and measurement detail for multiple use cases of trait-based research. We also provide computational tools in an R package to help users transfer their original data into the proposed scheme.

We build upon the structure proposed for EOL TraitBank [@parr16], which applies and expands terms of the Darwin Core standard [DWC; @wieczorek12]. We further expand the definitions of these terms  and add new key terms for the use case of trait data, particularly considering the work of @kattge11, and @garnier17, and drawing on the combined expertise of empirical biodiversity researchers (data providers), biodiversity synthesis researchers (data users), and biodiversity informatics researchers (data managers).

The vocabulary proposed will not capture the entire complexity of trait-based research in ecology and evolution. We invite researchers to comment on and contribute to the next iterations of this standard and develop own applications and ontologies.

@@

As a minimum consensus, trait datasets may be defined as follows: 
A trait dataset contains quantitative measurements or qualitative facts (i.e. trait values) about physical phenotypic characteristics of fitness, behaviour or performance (i.e. traits) observed on an individual organism or parts of an individual (i.e. an entity) that can be assigned to a biological taxon (i.e. a species or higher-level taxon) <!-- or larger organisational unit, e.g. a community -->.

Quantitative measurements are values obtained either by direct morphological or physiological measurement on single specimens (Fig. 1*a*), by aggregation of replicated measurements on multiple entities (Fig. 1*b*) or by estimation of means or ranges for the species as reported in literature or other published sources (e.g. databases, Fig. 1*c*). Qualitative facts are assignments of an entity to a categorical level, e.g. of a behavioural or life-history trait (Fig. 1.*d*). The entity or observation (i.e. the occurrence) to which the reported measurement or fact applies may differ in resolution -- depending on the scientific question -- and could be a subsample or bodypart, an individual specimen, an entire species or higher-level taxon.  
This definition refines the entity-quality model (EQ) as described by @garnier17, where a trait observation is phrased as 'an entity having a quality'. 
<!-- Cat: the difference [between quantitative measurements or qualitative facts] is not totally clear to me - is it the difference between field vs literature gathered data?  Flo: I see measurements as quantitative and facts as qualitative traits, which mostly aligns with field vs. literature, but not entirely. A differentiation that has rarely been specified in the trait-litereature it seems. Garnier et al 2017 just speak of "quality", which is assigned to an "entity", without specifiying it further. Nadja: I like this differentiation, to me this nicely reflects morphometric and literature traits (although its of course not exclusive). Maybe it helps if we add an example after the definition? --> <!--FDS: is this sufficiently defined now? the distinction between measured and literature data is spelled-out further below, when discussing the formats. Quantitative values can be taken from literature, though, like body sizes for example. -->

@@

Trait-datasets usually take different formats. For instance, if trait data have been collated at the species level from different literature sources or from expert knowledge, or as aggregated measurements collated from raw data, they usually are reported in a species$\times$trait matrix (a.k.a. wide-table format,  Fig. 2*a*), with a column of trait values for each trait recorded and a row for each species (or taxon) for which data were available. This format is usually reporting missing data as 'NA'. It may store additional information (e.g. on variation of means or literature source) in secondary colums or accompanying datasheets. 
The matrix format is widely used for the production of lookup-tables at the species level, which may either report qualitative facts (from literature or  expert knowledge) or aggregated quantitative traits (averages or ranges for this taxon). Quantitative traits at this resolution for instance may be used for the calculation of community weighted means or functional diversity metrics at the community level [@laliberte10]. Also at this level of information, missing entries, e.g. about the trait values or behaviour and functional role of a rarely described species, may be inferred from these tables combined with phylogenies ['gap filling', @swenson14] and using bayesian approaches [@schrodt15]. However, the tables do not allow to estimate within-species variation or assess the quality of the reported measurement or fact [@guralnik15].

Phenotypic variation of traits within species would usually be recorded per observation, i.e. each individual specimen would be recorded in a row, with the respective trait values stored in columns. In these occurrence data tables (Fig. 2*b*), each row is centered around a single physical occurrence of a species. Those data are common in investigations of evolutionary trade-offs and trait correlation, or of intra-specific variation along environmental gradients. Occurrence data would also capture phenotypic variation arising from morphotypes and sexual dimorphism. This format is the most intuitive for recording own empirical measurements and therefore is common for measured quantitative data and rarely found for reported qualitative facts obtained from the literature.

Computationally more effective and allowing for highest flexibility is the storage of traitdata in 'tidied' long-table formats [@wickham14; @parr15], where each row is reserved for a single measurement or fact of a specific trait, referenced to a single specimen  which is assigned to a specific taxon (i.e., one occurrence). This allows for storing repeated measurements on a single individual by linking the data of separate rows via a unique specimen ID. Also, multivariate trait measurements can be recorded in this format by linking multiple rows via a unique measurement identifier.  Longtable dataset purport multiple advantages for the data manipulation (e.g. filtering, subsetting and aggregating data), visualization (e.g. plot measured values by factor variable or taxon) and statistical modeling (e.g. ANOVA for testing difference of trait value by sex) [@wickham14]. They further draw from a defined set of columns, which simplifies merging of datasets.

The latter format thus provides the highest resolution and is ideal for storing raw data of trait measurements in huge databases and for linking them via the respective unique identifiers to additional information, such as occurrenceID or locationID. Each row of the dataset can be interpreted as a statement of an 'entity x having a qualitative/quantitative feature y'. 
To comply with the ETS, datasets must be reformatted to the longtable format (Fig. 2*c*).

@@

In the subsequent sections we propose a set of terms that describe the columns composing a trait dataset, starting with the ones required for our minimal definition of 'core' trait-datasets proposed above. To enable a better standardization of nomenclatures and values, we add columns that contain 'harmonized' entries, which are linked to semantic ontologies and globally valid identifiers. 
We furthermore provide extensions to add information on the level of the occurrence, the measurement or fact, and the entire dataset (i.e. metadata).

Additionally, a set of terms is proposed to facilitate the definition of traits in trait ontologies or a thesaurus of traits, consistent with semantic web standards (@ref, see below for further detail).

The online reference glossary of terms is found at https://ecologicaltraitdata.github.io/TraitDataStandard/ and should be referred to in the metadata of any dataset that employs the ETS as "Schneider et al. 2017 Ecological Traitdata Standard v1.0, DOI: XXXX.xxxx, URL: https://ecologicaltraitdata.github.io/TraitDataStandard/v1.0/". In addition to the versioned online reference, the dataset should also cite this publication for an explanation of the rationale.  
Individual terms should be referenced via their global Identifier, which is hosted by the GFBio Terminology Service and also can be accessed programmatically via an API (@ref). Where our glossary refines or duplicates existing terms from other ontologies for biological data, like the Glossary of EOL (http://eol.org/info/516) and Darwin Core (http://rs.tdwg.org/dwc/terms/), we indicate this in the fields 'refines' or 'identical', respectively.

@@

For these core values, we highly recommend to use unambiguous and self-explanatory vocabularies for trait names and taxa. However, to have the data compatible with own databases or analytical code, it might be necessary to use abbreviations or identifiers instead. It is indispensible to relate the user-defined vocabulary to a consensus standard of taxon names or or look-up tables of traits.

@@

Taxonomic standardisation, with respect to ongoing renamings and refinemenets, is a challenging issue in its own right. To enable compatibility across datasets, conventions on trait definitions and taxon names are inevitable. We consider the fact that most authors have their own schemes for standardisation and may refer to different scientific community standards by allowing for a double record of both user-specific and standardised entries [as practiced in TRY; @kattge11]. This duplication of data enables continuity on the author and data owner side and quality checking and comparability on the data user side. We therefore propose to add a second set of columns that contain standardised entries of the taxon name (`scientificNameStd`) which map synonyms to an accepted species name according to a published taxonomic ontology (e.g. GBIF Backbone Terminology) and links to it using a globally unique identifier (`taxonID`). The field `taxonID` therefore should contain an public,  globally unique Uniform Resource Identifier (URI) that resolves to the precise taxon definition at the time of the description or measurement (e.g. for *Bellis perennis*, the taxonID would be  'https://www.gbif.org/species/3117424'). 
Keeping a reference to a kingdom and class (using columns `kingdom`, `class` as described by DWC and suggested for EOL Traitbank) will avoid ambiguities with taxonomic homonyms. 
Additionally, it is best practice to keep a record of the reported taxon rank (in column `taxonRank`) for filtering purposes for traits recorded at the family or genus level. This information can be used for trait inference of higher taxonomic resolution using hierarchical probabilistic matrix factorisation [@schrodt15]. 
Alternatively to expanding the number of columns, these data could be contained in a separate data file which relates to the core dataset via the field `ScientificName`.

@@

A similar approach to standardised trait names and harmonised reported values in compliance with published ontologies is advised. Here, column `traitNameStd` should report the accepted trait name in an online ontology, while the value and unit are reported in columns `traitValueStd` and `traitUnitStd`.
Similar as for `taxonID`, the field `traitID` should refer to globally unique identifiers (URIs) of trait definitions or a measurement methodology.

<!-- The initiatives to standardise traits into consensus terminology are usually formed around a specific research question, methodology, or organism group and therefore vary in the traits that they consider important.  -->
To date, semantic ontologies exist for plant traits [e.g. TOP Plant Trait Ontology , @garnier17], soil invertebrates [T-SITA, @pey14], or biomedical traits of vertebrates [Vertebrate Trait Ontology, @park13] and more use cases (see OBO foundry for further examples). However, for most organism groups and research contexts we attest a deficit of public ontologies with publicly accessible URIs which would be used as `taxonID`.

@@

Relating the trait definition to methodological definitions on trait measurements help standardize the physical process of measurements. The Handbook for Measurement of traits [@moretti17] is a model example of such a methodological standardization approach.

Each trait definition may link to a broader or narrower term (`broaderTerm`, `narrowerTerm`), related terms (`relatedTerm`) or synonyms (`synonym`). For instance, the definition of 'femur length of first leg, left side' is narrower than 'femur length' which is narrower than 'leg trait' which is narrower than 'locomotion trait' (@ref semantic database methods). This extends the trait list into a semantic web resource, facilitates the classification of traits, and allows for cross-taxon comparative studies at the level of broader terms. (more detail on semantic web ontologies can be found in the online reference for the ETS at @refs...).

The existing thesauri and ontologies differ widely in terms of hierarchical depths and detail, as well as curation efforts and measures for peer-reviewed quality control. More compatibility of trait thesauri could greatly improve the possibilities for a comparison of trait data from different sources. 
We encourage hosting such semantic web resources developed for own projects as public resources, e.g. on OBO Foundry, the GFBio Terminology service, or on GitHub. We provide an example thesaurus, that has been developed to capture traits assessed within the German Biodiversity Exploratories project on Github ().

@@

<!-- 
(We conceptualize trait-datasets to  not include information about a species' interaction partners/associations of an organism, since they are -- to our understanding -- not within the scope of functional traits. However, such network data may be of interest to be combined with trait data. ) <!-- Malte: which can only be achieved if both types of datasets have been related to an ontology of species names. -->

@@

Beyond measurement units or taxon rank resolution, further information might be available that are not recorded in the core data table, but are related to the trait type, the taxon, the individual or specimen, or to the reported fact, measurement or sampling event.

To avoid redundancy and duplication of content, those information would usually be kept in seperate data sheets of a spreadsheet document. Examples are the place were a specimen has been sampled, the measurement method or statistical precision of the reported value, or the literature source from where a species value has been cited. A unique identifier would link to this other datasheet, encoding each individual occurrence of a species (`occurrenceID`), single measurements or reported facts (`measurementID`), locations of sampling (`locationID`) and sampling campaigns (`eventID`).

For reasons of reproducibility and openness, data should not be uploaded in proprietary spreadsheet formats (like '.xlsx') but rather in comma-separated text files ('.csv' or '.txt'). EOL TraitBank proposes the use of Darwin Core Archives [http://eol.org/info/structured_data_archives, DwC-A; @robertson09] to upload such related data along with the core trait dataset, which is also the accepted format for GBIF [@braak11]. These are .zip archives containing data table txt-files along with a descriptive metadata file (in .xml format). Detailled descriptions and tools for validation can be found on the website of EOL (http://eol.org/info/cp_archives) and GBIF (http://tools.gbif.org/dwca-assistant/).

Having separate data sheets linked by identifiers has the advantage of tidy datasets, avoiding duplication of information (@wickham14).
However, for standalone dataset publications on a hosting service without such file format recommendations (e.g. Figshare), data authors may want to avoid the expansion into multiple data tables and rather keep all columns in the same file, which is legitimate. 
It also may be safer to add such information directly as own columns of the data table to facilitate an analysis of cofactors and correlations further down the road. This way, if datasets of different source are merged, the information is readily available without the risk of breaking the reference to an external datasheet.
The trade-off is data-consistency and readability vs. avoidance of duplication.

Other cases, where key data columns would be placed in the same table as the core data are traits assessed on a higher level of organisation, e.g. microbial functional traits assessed at the community level taken from a soil sample. Here, location or measurement information are in the primary focus of the investigation. 
A general definition, whether a column describes asset data or is part of the central dataset is ill advised. Therefore, our glossary of terms should be used to describe the scientific data according to the study context.

The ETS provides two extensions of the namespace that should be used to describe these datasets:

@@

Many terms of this section refine terms of the Darwin Core Standard and their own [MeasurementOrFact](http://rs.tdwg.org/dwc/terms/index.htm#measureindex) and [Occurrence](http://rs.tdwg.org/dwc/terms/index.htm#Occurrence) extensions  and EOL TraitBank use of those terms (http://eol.org/info/structured_data_archives).

@@

@@

Scientific taxon names are matched automatically to the GBIF Backbone Taxonomy and added as duplicate columns by calling the function `standardize.taxonomy()`

The `standardize.traits()` function matches user-provided trait names onto public trait ontologies and links it to an existing trait ontology or a user-provided table of trait definitions, i.e. a thesaurus of traits. The same function harmonises trait values into target units and legit factor levels.

@@

Since the intention of the package is to facilitate upload of own data to hosting services but also simplify the access to published trait data, the package includes data objects that are compiled from datasets listed in our online spreadsheet that have been published in the public domain [@gossner15; @gossner15a; @ricklefs17; @vanderplas17]. 
To access data from public data-bases, the R-package 'traits' [@chamberlain17] contains functions to extract trait data via several open API interfaces including Birdlife, EOL TraitBank or BetyDB. The package  'TR8' provides similar access to plant traits from a list of databases (including LEDA, BiolFlor and Ellenberg Values)[@bocci15] and aggregates them into a species-trait matrix. However, none of these packages provides harmonization of traits into a unified scheme to facilitate further analysis.  
We provide a script (see online appendix and R-package vignette) to demonstrate the harmonization of these different data sources into our data standard.
